---
title: "Performing a typed analytical plan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performing a typed analytical plan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Render time: under 1 s under rmarkdown::render(); macOS arm64, R 4.6.1,
     one core. Two linear models and two rank correlations on 136 rows. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r library}
library(conductoR)
```

## The question

Does nitrogen rate affect maize yield? `agridat::hernandez.nitrogen` holds
136 plots from five Minnesota sites, four replicates each, at nitrogen rates
from 0 to 268 kg/ha. Two methods answer the question independently: the
slope of a linear regression of yield on rate, and Kendall's rank
correlation. A plan puts a power-aware gate on the rank lens, reconciles the
two lenses, and turns the consensus into a recommendation.

```{r data}
data("hernandez.nitrogen", package = "agridat")
d <- hernandez.nitrogen
table(d$site, d$rep)
```

## The tools

A tool is a function of `(inputs, data, node)` that returns an
`orchestraManifest::orchestra_manifest`. `lens_to_manifest()` lifts a
yes-or-no verdict into one; `decision_to_manifest()` does the same for a
recommendation. The tools are registered by name so the plan can be written
to a file.

```{r tools}
ols_lens <- function(inputs, data, node) {
  fit <- lm(yield ~ nitro, data = data)
  co  <- summary(fit)$coefficients["nitro", ]
  p   <- unname(co["Pr(>|t|)"])
  lens_to_manifest(
    headline = if (p < 0.05) "effect" else "no_effect",
    effect_present = p < 0.05, emitter = "ols", method = "ols:slope_t",
    metrics = list(p_value = p, slope = unname(co["Estimate"]),
                   effect_mean = unname(co["Estimate"]) * diff(range(data$nitro)),
                   noise_sd = sigma(fit), n = nrow(data)))
}

rank_lens <- function(inputs, data, node) {
  ct <- suppressWarnings(cor.test(data$nitro, data$yield, method = "kendall"))
  lens_to_manifest(
    headline = if (ct$p.value < 0.05) "effect" else "no_effect",
    effect_present = ct$p.value < 0.05, emitter = "rank", method = "rank:kendall",
    metrics = list(p_value = ct$p.value, tau = unname(ct$estimate),
                   n = nrow(data)))
}

recommend <- function(inputs, data, node) {
  consensus <- inputs$consensus@summary$headline
  action <- switch(consensus,
    causal_effect_corroborated = "apply_nitrogen",
    no_effect_corroborated     = "withhold_nitrogen",
    "insufficient_evidence")
  decision_to_manifest(action, emitter = "agronomy", method = "agronomy:rule",
                       metrics = list(consensus = consensus),
                       consumed = list(inputs$consensus))
}

register_tool("ols_lens", ols_lens)
register_tool("rank_lens", rank_lens)
register_tool("recommend", recommend)
```

## The plan

Five nodes. Each edge names the inferential target it expects from its
source; the gate reads the rank lens and abstains when a null verdict comes
from fewer than 25 plots or from a design that could not have resolved the
implied effect.

```{r plan}
score <- oap_plan(
  id = "nitrogen_response",
  meta = list(question = "Does nitrogen rate affect maize yield?"),
  nodes = list(
    node("lens_ols",  "ols_lens",  emits = "treatment_effects", role = "regression"),
    node("lens_rank", "rank_lens", emits = "treatment_effects", role = "rank"),
    node("gate", operator = "abstain_gate", emits = "decisions",
         consumes = list(edge("lens_rank", "treatment_effects")),
         params = list(n_min = 25L, z_floor = 2, mode = "flag")),
    node("consensus", operator = "triangulate", emits = "decisions",
         consumes = list(edge("lens_ols", "treatment_effects"),
                         edge("lens_rank", "treatment_effects"),
                         edge("gate", "decisions"))),
    node("decision", "recommend", emits = "decisions", role = "agronomy",
         consumes = list(edge("consensus", "decisions")))
  ))
plan_order(score)
```

## Scenario A: the full trial

```{r scenario_a}
run_a <- conduct(score, data = d)
run_a$terminal@summary$headline
run_a$bus$consensus@summary$metrics[c("agree", "n_lenses")]
```

Both lenses find the effect, the gate passes, and the consensus is
corroborated.

## Scenario B: one site, one replicate, three rates

```{r scenario_b}
small <- d[d$site == "S1" & d$rep == "R1" & d$nitro <= 67.2, ]
nrow(small)
run_b <- conduct(score, data = small)
run_b$bus$gate@summary$abstain_reason
run_b$terminal@summary$headline
```

With three plots the rank test cannot reject, and the gate refuses to let
that null through. In `flag` mode the consensus abstains and the
recommendation is `insufficient_evidence`. Rerouting instead drops the gated
lens and reconciles what survives.

```{r scenario_b_reroute}
gate <- plan_node(score, "gate")
gate@params$mode <- "reroute"
rerouted <- score
rerouted@nodes[[3]] <- gate
run_b2 <- conduct(rerouted, data = small, verbose = FALSE)
run_b2$bus$consensus@summary$metrics[c("rerouted", "n_lenses")]
run_b2$terminal@summary$headline
```

One lens on three plots is not a corroborated effect, so the recommendation
is still to wait for evidence.

## The plan as a file

Because every compute node names its tool, the plan writes to YAML and
reads back to the same run.

```{r yaml}
f <- tempfile(fileext = ".oap.yaml")
plan_to_yaml(score, f)
cat(readLines(f, n = 12), sep = "\n")
again <- plan_from_yaml(f)
run_c <- conduct(again, data = d, verbose = FALSE)
identical(run_c$terminal@run_id, run_a$terminal@run_id)
```

## A mistyped edge

An edge that declares the wrong contract is refused before the downstream
node runs. Here the decision node is told to expect `parameters` from the
consensus, which emits `decisions`.

```{r violation}
bad <- oap_plan(id = "mistyped", nodes = list(
  node("lens_ols", "ols_lens", emits = "treatment_effects"),
  node("decision", "recommend", emits = "decisions",
       consumes = list(edge("lens_ols", "parameters")))
))
res <- try(conduct(bad, data = d, verbose = FALSE), silent = TRUE)
cat(conditionMessage(attr(res, "condition")))
```

## Provenance and a targets pipeline

Giving `conduct()` an `out_dir` writes `run.rds`, `run_card.md` and
`provenance.md`, the last carrying the chain of payload hashes in execution
order. `run_provenance()` returns the same table in the session.

```{r provenance}
run_provenance(run_a)[, c("node", "operator", "emits", "abstained")]
```

The same plan compiles to a `_targets.R` in which each node is one target
calling `conduct_step()`, so a long-running plan inherits caching from the
'targets' package.

```{r targets, eval = FALSE}
compile_to_targets(score, "_targets.R", preamble = c(
  "library(conductoR)",
  "source('tools.R')",
  ".score <- plan_from_yaml('nitrogen_response.oap.yaml')",
  ".data <- agridat::hernandez.nitrogen"))
targets::tar_make()
targets::tar_read_raw(terminal_target(score))
```
